Testing the motor and cognitive foundations of Paleolithic social transmission

### 2021-08-18 manuscript/manuscript.Rmd
supplementary data created.

---
title: Testing the motor and cognitive foundations of Paleolithic social transmission
author: 
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
- Megan Beney Kilgore^[Department of Anthropology, Emory University, Atlanta, GA, USA; megan.elizabeth.beney@emory.edu]
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
abstract: "Stone tools provide key evidence of human cognitive evolution but remain difficult to interpret. Toolmaking skill-learning in particular has been understudied even though: 1) the most salient cognitive demands of toolmaking should occur during learning, and 2) variation in learning aptitude would have provided the raw material for any past selection acting on tool making ability. However, we actually know very little about the cognitive prerequisites of learning under different information transmission conditions that may have prevailed during the Paleolithic. This paper presents results from a pilot experimental study to trial new experimental methods for studying the effect of learning conditions and individual differences on Oldowan flake-tool making skill acquisition. We trained 23 participants for 2 hours to make stone flakes under two different instructional conditions (observation only vs. direct active teaching) employing appropriate raw materials, practice time, and real human interaction. Participant performance was evaluated through analysis of the stone artifacts produced. Performance was compared both across experimental groups and with respect to individual participant differences in grip strength, motor accuracy, and cognitive function measured for the study. Our results show aptitude to be associated with fluid intelligence in a verbally instructed group and with a tendency to use social information in an observation-only group. These results have implications for debates surrounding the cumulative nature of human culture, the relative contributions of knowledge and know-how for stone tool making, and the role of evolved psychological mechanisms in “high fidelity” transmission of information, particularly through imitation and teaching.\\par \\textbf{Keywords:} Oldowan; Stone toolmaking; Social learning; Individual variation; Cognitive aptitudes; Motor skills"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

"*The many slight differences... being observed in the individuals of the same species inhabiting the same confined locality, may be called individual differences... These individual differences are of the highest importance to us, for they are often inherited ... and they thus afford materials for natural selection to act on and accumulate...*" [@darwin1859, Chapter 2]

@@

"*A creature that learns to make tools to a complex pre-existing pattern...must have the kind of abstracting mind that would be of high selective value in facilitating the development of the ability to communicate such skills by the necessary verbal acts.*"[@montagu1976: 267]

@@

*Such undertakings -- based on raw material which is never standard, and with gestures of percussion that are never perfectly delivered -- cannot be reduced to an elementary repetition of gestures...the realization of elaborate knapping activities necessitates a critical monitoring of the situation and of the decisions adopted all through the process. [@pelegrin1990: 117]*

>>>>

---
title: Testing the motor and cognitive foundations of Paleolithic social transmission
author: 
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
- Megan Beney Kilgore^[Department of Anthropology, Emory University, Atlanta, GA, USA; megan.elizabeth.beney@emory.edu]
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Aditi Majoe^[Indepedent Researcher; am2752@cantab.ac.uk]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
abstract: "Stone tools provide key evidence of human cognitive evolution but remain difficult to interpret. Toolmaking skill-learning in particular has been understudied even though: 1) the most salient cognitive demands of toolmaking should occur during learning, and 2) variation in learning aptitude would have provided the raw material for any past selection acting on tool making ability. However, we actually know very little about the cognitive prerequisites of learning under different information transmission conditions that may have prevailed during the Paleolithic. This paper presents results from a pilot experimental study to trial new experimental methods for studying the effect of learning conditions and individual differences on Oldowan flake-tool making skill acquisition. We trained 23 participants for 2 hours to make stone flakes under two different instructional conditions (observation only vs. direct active teaching) employing appropriate raw materials, practice time, and real human interaction. Participant performance was evaluated through analysis of the stone artifacts produced. Performance was compared both across experimental groups and with respect to individual participant differences in grip strength, motor accuracy, and cognitive function measured for the study. Our results show aptitude to be associated with fluid intelligence in a verbally instructed group and with a tendency to use social information in an observation-only group. These results have implications for debates surrounding the cumulative nature of human culture, the relative contributions of knowledge and know-how for stone tool making, and the role of evolved psychological mechanisms in “high fidelity” transmission of information, particularly through imitation and teaching.\\par \\textbf{Keywords:} Oldowan; Stone toolmaking; Social learning; Individual variation; Cognitive aptitudes; Motor skills"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

|   "*The many slight differences... being observed in the individuals of the same species inhabiting the same confined locality, may be called individual differences... These individual differences are of the highest importance to us, for they are often inherited ... and they thus afford materials for natural selection to act on and accumulate...*" [@darwin1859, Chapter 2]

@@

|   "*A creature that learns to make tools to a complex pre-existing pattern...must have the kind of abstracting mind that would be of high selective value in facilitating the development of the ability to communicate such skills by the necessary verbal acts.*"[@montagu1976: 267]

@@

|   "*Such undertakings -- based on raw material which is never standard, and with gestures of percussion that are never perfectly delivered -- cannot be reduced to an elementary repetition of gestures...the realization of elaborate knapping activities necessitates a critical monitoring of the situation and of the decisions adopted all through the process." [@pelegrin1990: 117]*
